---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: true  
---

**Reading in the data**  
To begin with, you must download the data set from [this website](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) and save it into your working directory. Assign the path of your directory to the variable Rep_Research. After you load the data, the only thing you need to fix is the class of the date variable

```{r loaddata, echo=TRUE}
Rep_Research <- "/Users/sommpd10/datasciencecoursera/Reproducible-Research"
setwd(Rep_Research)
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
file <- "activity"
download.file(url, file, method = "curl")
unzip(file, exdir = Rep_Research)
data <- read.csv("activity.csv", header=TRUE)
data$date <- as.Date(data$date, format = "%Y-%m-%d")
```

**Question 1**  
For this question, you must calculate the total number of steps taken each day and make a histogram showing the distribution of the number of steps taken every day.

First, make a table:
``` {r q1table, echo=TRUE}
Daily_Sums <- aggregate(data$steps, by = list(data$date), FUN = "sum")
colnames(Daily_Sums) <- c("Date", "Total_Steps")
head(Daily_Sums)
```

Next, make the plot:
``` {r q1plot, echo=TRUE}
library(ggplot2)
ggplot(Daily_Sums, aes(x = Total_Steps)) + geom_histogram(binwidth = 1000, fill="white", color="black") + labs(title="Distribution of Total Steps per Day", x = "Total Steps per Day")
```

Lastly, calculate the mean and median:
```{r q1medmean, echo=TRUE}
Q1mean <- mean(Daily_Sums$Total_Steps, na.rm=TRUE)
Q1median <- median(Daily_Sums$Total_Steps, na.rm=TRUE)
```

The mean is `r Q1mean` and the median is `r Q1median`.

**Question 2**  
For this question, you must explore daily trends of steps taken during 5 minute intervals throughout the day.

First, make a table (and note that to calculate means, you must drop all missing values):
``` {r q2table, echo=TRUE}
data_noNAs <- data[!is.na(data$steps), ]
Interval_Means <- aggregate(data_noNAs$steps, by = list(data_noNAs$interval), FUN = "mean")
colnames(Interval_Means) <- c("Time_Interval", "Mean_Steps")
head(Interval_Means)
```

Next, make the plot:
``` {r q2plot, echo=TRUE}
library(ggplot2)
g <- ggplot(Interval_Means, aes(x = Time_Interval, y = Mean_Steps))
g + geom_line() + labs(title="Average Number of Steps \n During 5 Minute Intervals Throughout the Day", x = "Time Interval", y = "Average Number of Steps")
```

Lastly, find the period during the day with the largest average number of steps:
```{r q2max, echo=TRUE}
Max_Interval <- Interval_Means[which.max(Interval_Means[, 2]), 1]
Max_Value <- Interval_Means[which.max(Interval_Means[, 2]), 2]
```

Maximum number of steps during a 5 minute period is `r Max_Value` and this takes place during the interval beginning at `r Max_Interval`.

**Question 3**  
For this question, you must impute the missing values for number of steps during a given time interval and then recalculate the distribution of total number of steps taken every day.

First, calculate the number of rows with missing data:
```{r q3missingnumber, echo=TRUE}
number_missingq3 <- (nrow(data) - sum(complete.cases(data)))
```
We find that `r number_missingq3` rows have missing data.

Next, I impute missing values for number of steps during a given time interval with the average number of steps during that interval. This is a three step process: calculate average number of steps during every interval (calculated above), add a column with the average number of steps during each interval, replace NAs with the average for that given interval.
```{r q3replacemissing, echo=TRUE}
data_NAs_Imputed <- merge(data, Interval_Means, by.x = "interval", by.y = "Time_Interval", all.x = TRUE)
data_NAs_Imputed$steps[is.na(data_NAs_Imputed$steps)] <- data_NAs_Imputed$Mean_Steps[is.na(data_NAs_Imputed$steps)]
```
To ensure that missing values are replaced, calculate the number of rows with missing values:
```{r q3missingnumberb, echo=TRUE}
number_missingq3b <- (nrow(data_NAs_Imputed$steps) - sum(complete.cases(data_NAs_Imputed$steps)))
number_missingq3b
```

The next step is to recalculate the total number of steps taken every day using the new data set with imputed values:
```{r q3table, echo=TRUE}
Daily_Sums_Imputed <- aggregate(data_NAs_Imputed$steps, by = list(data_NAs_Imputed$date), FUN = "sum")
colnames(Daily_Sums_Imputed) <- c("Date", "Total_Steps")
head(Daily_Sums_Imputed)
```

The last step is to calculate the new mean and median value for total number of steps taken per day as well as to plot a new histogram:
```{r q3newplot, echo=TRUE}
Mean_Imputed <- mean(Daily_Sums_Imputed$Total_Steps, na.rm=TRUE)
Median_Imputed <- median(Daily_Sums_Imputed$Total_Steps, na.rm=TRUE)

ggplot(Daily_Sums_Imputed, aes(x = Total_Steps)) + geom_histogram(binwidth = 1000, fill="white", color="black") + labs(title="Distribution of Total Steps per Day \n Using Date Set with Imputed Values", x = "Total Steps per Day")
```

Below, we can compare the mean and median values from the two data sets:   
- Without Imputation: mean = `r Q1mean` and median = `r Q1median`  
- With Imputation: mean = `r Mean_Imputed` and median = `r Median_Imputed`

We can see that both the mean and median remain the same after imputation.

**Question 4**  
For this question, you must compare the number of steps taken during 5 minute intervals during days in the week versus days in the weekend.

First, you must create a new variable that indicates whether the date is a weekday or weekend and make sure to use the data set with imputed values:
```{r q4weekdayfactor, echo=TRUE}
data_NAs_Imputed$Weekday <- ifelse(weekdays(data_NAs_Imputed$date) != "Saturday" & weekdays(data_NAs_Imputed$date) != "Sunday", "Weekday", "Weekend")
data_NAs_Imputed$Weekday <- as.factor(data_NAs_Imputed$Weekday)
```

The next step is to create a table that calculates the average number of steps taken during 5 minute intervals for weekdays and for weekend days:
```{r q4frequencytable, echo=TRUE}
Interval_Means_Weekday <- aggregate(data_NAs_Imputed$steps, by = list(data_NAs_Imputed$Weekday, data_NAs_Imputed$interval), FUN="mean")
names(Interval_Means_Weekday) <- c("Weekday", "Interval", "Mean_Steps")
head(Interval_Means_Weekday)
```

The last step is to plot daily trends of steps taken throughout the day during weekdays and weekend days:
```{r q4plot, echo=TRUE}
library(ggplot2)
g <- ggplot(Interval_Means_Weekday, aes(x = Interval, y = Mean_Steps))
g + geom_line() + facet_grid(Weekday ~ .) + labs(title="Average Number of Steps \n During 5 Minute Intervals Throughout the Day", x = "Time Interval", y = "Average Number of Steps")
```